---
title: Plots With Persephone3
description: >
  Overview about the different chart types that are available with the
  persephone3 package.
---

Overview about the different chart types that are available with the
`{persephone3}` package.
Note that these charts are all interactive and based on `{highcharter}`.
If you are interested in print graphics, `{ggdemetra3}` might be more relevant. <- TO DO: Implement ggplot print graphics from old persephone without interactive option

## TODO

* [spectral plots](https://statistikat.github.io/persephone/reference/plotSpectrum.html)
* [residual plots](https://statistikat.github.io/persephone/reference/plotResiduals.html)
* [calendar effects](https://statistikat.github.io/persephone/reference/plotSeasIrrCal.html)
* [si ratios](https://statistikat.github.io/persephone/reference/plotSiRatios.html)


```{r, include=FALSE}
devtools::load_all()
```


```{r}
library(persephone3)
x <- perX13(AirPassengers)
x$run()
hchart(x)
plotSeasIrrCal(x)
```

```{r}
y <- perTramo(nottem)
y$run()
hchart(y) %>% 
  hc_title(text = "Average Monthly Temperatures at Nottingham")
plotSeasIrrCal(y)
```
