---
title: "Agentic seasonal adjustment with perAgent"
description: >
  How `perAgent()` iterates through conservative seasonal-adjustment updates,
  how to read the agent history, and where `perAgentOpenAI()` fits into the
  workflow.
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Agentic seasonal adjustment with perAgent}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

`perAgent()` is a conservative orchestration layer on top of the regular
`perX13()` and `perTramo()` workflows. Instead of jumping directly to one
hand-crafted specification, it starts from a standard template, checks the
diagnostics, proposes a small set of explainable updates, and keeps the full
search history.

This article shows that workflow on an aggregate production-in-industry series
built from `pi_unadj`, following the same data family used in
`vignettes/resources/slides_ntts.Rmd.src`. It also shows how to present the
agent's updates in a readable way and where `perAgentOpenAI()` fits in.

## How the agent works

At a high level, `perAgent()` follows four steps:

1. It runs pre-tests for seasonality.
2. In `method = "auto"` mode it compares an initial X-13ARIMA-SEATS run with an
   initial TRAMO-SEATS run.
3. It creates a short list of conservative candidate actions such as enabling a
   log transform, adding working-day regressors, trying holiday-adjusted
   trading-day variants based on the built-in Austrian holiday calendar,
   enabling Easter, tightening automatic outlier detection, freezing the
   current ARIMA model, or switching the seasonal-adjustment family.
4. It re-runs those candidates, accepts only the ones that improve the score
   enough, and stores the whole search path in `result$history`.

The score is driven by diagnostics that matter for official-statistics style
workflows: residual seasonality, residual trading-day effects, the Q statistic,
and a few stability-oriented penalties.

## Build the aggregate series

We follow the same setup as in the NTTS slides and start from `pi_unadj` plus
the EA-19 weights. The stored aggregate columns in `pi_unadj` are empty, so we
reconstruct the aggregate from the country series and keep the common sample
where all selected components are observed.

```{r}
library(persephone3)

data(pi_unadj)
data(weights_pi_ea19)

weighted_aggregate <- function(x, weights) {
  w <- weights$weight[match(colnames(x), weights$country)]
  values <- apply(x, 1, stats::weighted.mean, w = w)
  stats::ts(values, start = stats::start(x), frequency = stats::frequency(x))
}

ea19_countries <- weights_pi_ea19$country
ea19_panel <- pi_unadj[, ea19_countries]
ea19_panel <- ea19_panel[, colSums(is.na(ea19_panel)) < nrow(ea19_panel)]
first_complete <- which(stats::complete.cases(ea19_panel))[1]
ea19_panel <- stats::window(ea19_panel, start = stats::time(ea19_panel)[first_complete])
ea19_weights <- weights_pi_ea19[match(colnames(ea19_panel), weights_pi_ea19$country), ]
ea19_aggregate <- weighted_aggregate(ea19_panel, ea19_weights)

c(
  start = paste(stats::start(ea19_aggregate), collapse = ", "),
  end = paste(stats::end(ea19_aggregate), collapse = ", "),
  frequency = stats::frequency(ea19_aggregate),
  components = ncol(ea19_panel)
)
```

For context, the next figure compares the unadjusted weighted aggregate with the
final adjusted series chosen by the agent.

```{r run-agent}
agent_fit <- perAgent(ea19_aggregate, method = "auto", max_iter = 4)
agent_fit
```

```{r plot-final-series}
plot_original <- stats::window(
  ea19_aggregate,
  end = stats::end(agent_fit$final_model$adjusted)
)

series_plot <- data.frame(
  date = zoo::as.Date(zoo::as.yearmon(stats::time(plot_original))),
  original = as.numeric(plot_original),
  adjusted = as.numeric(agent_fit$final_model$adjusted)
)

ggplot2::ggplot(series_plot, ggplot2::aes(date)) +
  ggplot2::geom_line(
    ggplot2::aes(y = original, color = "Original aggregate"),
    linewidth = 0.5,
    alpha = 0.75
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = adjusted, color = "Final adjusted series"),
    linewidth = 0.6
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "Original aggregate" = "#6c757d",
      "Final adjusted series" = "#005f73"
    )
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Index",
    color = NULL,
    title = "Weighted EA-19 aggregate from pi_unadj"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

In this run the agent ends on `r agent_fit$selected_method` with a final score
of `r round(agent_fit$final_score, 2)`.

The dedicated plot method gives a compact view of the whole search: the
original aggregate, the final adjusted series, and the earlier successful runs
in the background.

```{r plot-agent-search}
plot(agent_fit)
```

## Present the accepted updates clearly

The most useful part of a `perAgent()` result is `history`. Each row is one
evaluated candidate, and `accepted = TRUE` marks the path that actually became
the current best model.

```{r history-helpers}
format_value <- function(x) {
  paste(as.character(x), collapse = ", ")
}

format_speclist <- function(x) {
  if (length(x) == 0) {
    return("default template")
  }

  paste(
    sprintf("%s = %s", names(x), vapply(x, format_value, character(1))),
    collapse = "; "
  )
}

format_num <- function(x, digits = 3) {
  ifelse(is.na(x), "NA", format(round(x, digits), nsmall = digits))
}

accepted_history <- agent_fit$history[agent_fit$history$accepted, ]
accepted_updates <- data.frame(
  round = accepted_history$iteration,
  method = accepted_history$method,
  action = accepted_history$action,
  score = round(accepted_history$score, 2),
  score_gain = c(NA, round(diff(accepted_history$score), 2)),
  residual_td_p = format_num(accepted_history$residual_td_pvalue),
  q_stat = format_num(accepted_history$q_stat),
  specification = vapply(accepted_history$speclist, format_speclist, character(1)),
  rationale = accepted_history$rationale,
  check.names = FALSE
)

knitr::kable(
  accepted_updates,
  caption = "Accepted path through the agent search"
)
```

That table is usually the clearest summary for a report: it shows the initial
choice, the accepted updates, the cumulative specification at each accepted
step, and the size of each score improvement.

## Show the full search history

For debugging or method-development work, it is often useful to keep the full
candidate log instead of only the accepted path.

```{r full-history}
history_display <- data.frame(
  step = seq_len(nrow(agent_fit$history)),
  round = agent_fit$history$iteration,
  status = ifelse(agent_fit$history$accepted, "accepted", "considered"),
  method = agent_fit$history$method,
  action = agent_fit$history$action,
  score = round(agent_fit$history$score, 2),
  residual_f_p = format_num(agent_fit$history$residual_seasonal_f_pvalue),
  residual_qs_p = format_num(agent_fit$history$residual_seasonal_qs_pvalue),
  residual_td_p = format_num(agent_fit$history$residual_td_pvalue),
  q_stat = format_num(agent_fit$history$q_stat),
  check.names = FALSE
)

knitr::kable(
  history_display,
  caption = "Full agent evaluation log"
)
```

```{r score-trajectory}
plot_history <- transform(
  agent_fit$history,
  step = seq_len(nrow(agent_fit$history)),
  status = ifelse(agent_fit$history$accepted, "accepted", "considered")
)

accepted_points <- plot_history[plot_history$accepted, ]

ggplot2::ggplot(plot_history, ggplot2::aes(step, score)) +
  ggplot2::geom_line(color = "#c0cad5", linewidth = 0.5) +
  ggplot2::geom_point(
    ggplot2::aes(color = status, shape = method),
    size = 2.8
  ) +
  ggplot2::geom_text(
    data = accepted_points,
    ggplot2::aes(label = action),
    vjust = -0.8,
    size = 3,
    check_overlap = TRUE
  ) +
  ggplot2::scale_color_manual(
    values = c(accepted = "#005f73", considered = "#94a1b2")
  ) +
  ggplot2::labs(
    x = "Candidate evaluation",
    y = "Agent score",
    color = NULL,
    shape = "Method",
    title = "Score trajectory across the search"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

Together, the accepted-path table and the full search log answer slightly
different questions:

- Use the accepted-path table when you want a concise narrative of what the
  agent changed.
- Use the full log when you want to understand which alternatives were
  considered and why they were not kept.

## Where perAgentOpenAI fits

`perAgentOpenAI()` does not replace the rule-based search. Instead, it creates a
small OpenAI-backed `llm_decider` that sees the current diagnostics, the search
history, and the candidate actions already generated by `perAgent()`.

That decider can then prioritise which candidate should be tried first. The
underlying safety rail stays the same: `perAgent()` still evaluates the
candidate, scores it, and accepts it only if it improves the diagnostics enough.

```{r openai-example, eval=FALSE}
decider <- perAgentOpenAI(
  model = "gpt-4.1-mini",
  system_prompt = paste(
    "Prefer updates that remove residual trading-day effects first.",
    "Only switch method when the diagnostics clearly improve."
  )
)

agent_fit_llm <- perAgent(
  ea19_aggregate,
  method = "auto",
  max_iter = 4,
  llm_decider = decider
)

agent_fit_llm$llm_used
agent_fit_llm$history[, c("iteration", "method", "action", "score", "accepted")]
```

In practice:

- `perAgentOpenAI()` is most useful when several candidate actions look
  reasonable and you want a language-model nudge on the order in which they are
  explored.
- It requires the suggested package `ellmer` plus OpenAI credentials, typically
  via `OPENAI_API_KEY`.
- It is still a prioritisation layer, not a free-form model builder.

If you want to keep the final production specification fully reproducible, you
can also ask the agent to freeze the winning model after the search:

```{r fixed-final-model, eval=FALSE}
agent_fit_fixed <- perAgent(
  ea19_aggregate,
  method = "auto",
  max_iter = 4,
  fix_outliers = TRUE,
  fix_model = TRUE
)
```
