---
title: "Hierarchical time series - construction and weights"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Hierarchical time series - construction and weights}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "95%"
)
```

Hierarchical time series are time series that can be divided into sub-series.
In `persephone3`, such objects are created similar to a nested list. The
following example creates a series with European aggregates based on export data from different European countries.

First we load the `Exports` data from `rjd3toolkit`:
```{r}
library(persephone3)

data(Exports, package = "rjd3toolkit")
```

For all countries' time series we now want to generate a persephone object with Tramo as method.

```{r}
# EU countries only (27 member states, excluding UK and non-EU countries)
eu_countries <- c("France", "Luxembourg", "Netherlands", "Germany", "Italy",
                  "Ireland", "Denmark", "Greece", "Portugal",
                  "Spain", "Sweden", "Finland", "Austria", "Malta", "Estonia",
                  "Latvia", "Lithuania", "Poland", "Czech.Republic", "Slovakia",
                  "Hungary", "Slovenia", "Cyprus", "Bulgaria", "Romania", "Croatia")

ts_28 <- lapply(Exports[eu_countries], perTramo)
```

We generate a additional layer based on a certain regional grouping of the countries.

```{r}
# Define regional groupings using the actual country names from the dataset
south_eastern_europe <- c("Greece", "Cyprus", "Bulgaria", "Romania")
northern_europe <- c("Denmark", "Finland", "Sweden", "Ireland")
central_europe <- c("Croatia", "Germany", "Czech.Republic", "Slovakia", "Hungary",
                    "Austria", "Slovenia", "Luxembourg", "Estonia", "Lithuania",
                    "Latvia", "Poland")
southern_europe <- c("Italy", "Spain", "Portugal", "Malta")
western_europe <- c("France", "Netherlands")
```

The list argument can be used to create a hierachical persephone object based on a list of persephone objects.

```{r}
hts_see <- perHts(list = ts_28[south_eastern_europe])
hts_ne <- perHts(list = ts_28[northern_europe])
hts_ce <- perHts(list = ts_28[central_europe])
hts_se <- perHts(list = ts_28[southern_europe])
hts_we <- perHts(list = ts_28[western_europe])
```


An additional layer is only the nothern part and southern part
```{r}
hts_np <- perHts(norhtern_europe = hts_ne,
                 western_europe = hts_we,
                 central_europe = hts_ce
)
hts_sp <- perHts(southern_europe = hts_se,
                 south_eastern_europe = hts_see
)
```

Finally we aggregate them to the whole of EU.
```{r}
ht_europe <- perHts(north_part = hts_np,
                 south_part = hts_sp
)
ht_europe
```

Now we make the first run:
```{r}
ht_europe$run()
ht_europe
```


# Weights

Weights can be used to customize the aggregation step, so the hierachical time series is not just the sum of its components, but the weighted sum. This can be especially helpful if the timeseries are indices.

## Time-invariant weights

The weights parameter will be set as numeric vector with length equal to the number of components.

Almost the same call, but with weights, one for each component.
```{r}
hts_see <- perHts(list = ts_28[south_eastern_europe], weights=c(1,2,3,4))
hts_ne <- perHts(list = ts_28[northern_europe], weights = c(5,6,7,8)) # 4 countries: Denmark, Finland, Sweden, Ireland
hts_ce <- perHts(list = ts_28[central_europe], weights = rep(1,12))
hts_se <- perHts(list = ts_28[southern_europe], weights = c(8,9,10,10.5))
hts_we <- perHts(list = ts_28[western_europe], weights = c(11,12)) # 2 countries: France, Netherlands
```

In additional aggregation steps, the weight is automatically summed up. So e.g. the weight for central europe is 12, the sum of 12 times 1.
```{r}
hts_np <- perHts(northern_europe = hts_ne,
                 western_europe = hts_we,
                 central_europe = hts_ce
)
hts_sp <- perHts(southern_europe = hts_se,
                 south_eastern_europe = hts_see
)
knitr::kable(head(hts_np$weights))
```


```{r}
ht_europe <- perHts(north_part = hts_np,
                 south_part = hts_sp
)

knitr::kable(head(ht_europe$weights))
```


## Time-variant weights

In many scenarios weights are not fixed over time, but change, e.g. if the turnover is used to aggregate a certain production index. In these scenarios weights can be provided as lists of time series or as a mts object.

We are starting with weights equal to 1 and then only change the weight of Germany over time.

```{r}
# Convert list to mts (multi-time series) object
weights_eu <- do.call(cbind, Exports[eu_countries]) * 0 + 1
weights_eu[is.na(weights_eu)] <- 1
weights_eu[,"Germany"] <- weights_eu[,"Germany"] + rlnorm(nrow(weights_eu))
```


We can now call the `perHts` function with the mts as input.
```{r}
hts_see <- perHts(list = ts_28[south_eastern_europe],
                   weights=weights_eu[,south_eastern_europe])
hts_ne <- perHts(list = ts_28[northern_europe],
                  weights=weights_eu[,northern_europe])
hts_ce <- perHts(list = ts_28[central_europe],
                  weights=weights_eu[,central_europe])
hts_se <- perHts(list = ts_28[southern_europe],
                  weights=weights_eu[,southern_europe])
hts_we <- perHts(list = ts_28[western_europe],
                  weights=weights_eu[,western_europe])
```

And then we can proceed as before.

```{r}
hts_np <- perHts(northern_europe = hts_ne,
                 western_europe = hts_we,
                 central_europe = hts_ce
)
hts_sp <- perHts(southern_europe = hts_se,
                 south_eastern_europe = hts_see
)
knitr::kable(head(hts_np$weights))
```


```{r}
ht_europe <- perHts(north_part = hts_np,
                 south_part = hts_sp
)

knitr::kable(head(ht_europe$weights))
```

# Next Steps

- See `vignette("persephone3")` for basic usage and template options
- See `vignette("persephone-plotting")` for visualization methods for hierarchical objects
- Check `?perHts` for complete documentation on hierarchical time series

# Repository

Source code and issue tracker: [https://github.com/statistikat/persephone3](https://github.com/statistikat/persephone3)
