---
title: "Hierarchical time series with persephone3"
description: >
  Working with hierarchical time series: creating, adjusting, and iterating
  over multiple levels of aggregation.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical time series with persephone3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{persephone3}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/persephone-hierarchical-",
  out.width = "100%"
)
```

# Introduction

Hierarchical time series are time series that can be divided into sub-series with different levels of aggregation. In `persephone3`, such objects are created similar to a nested list structure.

The following example creates a hierarchical object with two subseries `a` and `b`:

```{r}
library(persephone3)

hts <- perHts(
  a = perX13(AirPassengers),
  b = perTramo(AirPassengers)
)
hts
```

## Running the Model

When the `run()` method of `hts` is invoked, the following adjustments are performed:

1. The aggregate series `a+b` is adjusted and the results are saved in `hts$output`
2. `a` is adjusted using X-13ARIMA-SEATS
3. `b` is adjusted using TRAMO/SEATS

The results can be viewed as follows:

```{r}
display_result <- function(ts) {
  window(ts, end = c(1949, 6))
}

hts$run()

# Direct adjustment (seasonally adjusted aggregate series)
display_result(hts$adjusted)

# Seasonally adjusted component series "a"
display_result(hts$components$a$adjusted)

# Indirect adjustment (sum of seasonally adjusted components)
display_result(hts$adjustedIndirect)
```

The last line displays the results of an indirect adjustment, which amounts to `a$adjusted` + `b$adjusted` in this case.

## Multiple Levels of Hierarchies

For more complex structures with multiple hierarchy levels, `perHts` calls can be nested:

```{r}
hts_nested <- perHts(
  hts = hts,
  hts_child = perHts(
    a = perX13(AirPassengers)
  ),
  b = perTramo(AirPassengers)
)

hts_nested
```

As you can see, the structure of `hts` is available within this nested object.

## Iterating Over the Hierarchy Tree

To iterate over all components of a `perHts` object, the `iterate()` method can be used. It behaves similarly to `lapply()` and allows you to apply custom functions to each node in the hierarchy:

```{r, warning=FALSE}
# Run the model first
hts_nested$run()

# Apply custom function to get mean and standard deviation
hts_nested$iterate(asTable = TRUE, function(obj) {
  list(
    mean = mean(obj$adjusted),
    sd = sd(obj$adjusted)
  )
})
```

## Setting Options Recursively

When creating a `perHts` object, you can pass specification parameters that will be applied to the aggregate series. With `iterate = TRUE` (the default), these parameters are also applied to all components recursively:

```{r, warning=FALSE}
# Create hierarchical object with custom parameters applied to all nodes
hts_custom <- perHts(
  a = perX13(AirPassengers),
  b = perTramo(AirPassengers),
  template = "rsa1",
  automdl.enabled = FALSE
)

hts_custom$run()
hts_custom
```

The `automdl.enabled = FALSE` setting was applied to both the aggregate series and both components (`a` and `b`). You can verify this by inspecting the specifications:

```{r}
# Check aggregate spec
hts_custom$paramsDirect$automdl.enabled

# Check component a spec
hts_custom$components$a$paramsDirect$automdl.enabled

# Check component b spec
hts_custom$components$b$paramsDirect$automdl.enabled
```

To modify individual components after creation, access them directly:

```{r, warning=FALSE}
# Re-enable automdl for component a only
hts_custom$components$a$updateParams(automdl.enabled = TRUE)
hts_custom$components$a$paramsDirect$automdl.enabled
```

# Next Steps

- See `vignette("persephone3")` for basic usage and template options
- See `vignette("persephone-plotting")` for visualization methods for hierarchical objects
- Check `?perHts` for complete documentation on hierarchical time series

# Repository

Source code and issue tracker: [https://github.com/statistikat/persephone3](https://github.com/statistikat/persephone3)
