---
title: "Plotting with persephone3"
description: >
  Interactive visualizations for seasonal adjustment results using dygraphs,
  plotly, and highcharter.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting with persephone3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{persephone3}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/persephone-plotting-",
  out.width = "100%"
)
```

# Introduction

`persephone3` provides multiple ways to visualize seasonal adjustment results. The plotting functions return interactive HTML widgets using `dygraphs`, `plotly`, and `highcharter`.

**Note:** The appearance of plots depends on whether `run()` has been called. Before running, only the raw time series is available. After `run()`, additional components (seasonally adjusted, trend-cycle, forecasts) become accessible.

```{r setup-example, message = FALSE, warning = FALSE}
library(persephone3)

# Create seasonal adjustment object
obj <- perX13(AirPassengers, "rsa1")
```

# Default Plot Method

The default `plot()` method creates an interactive time series plot using `dygraphs`. You can zoom into specific periods and hover to see exact values.

```{r default-plot}
# Plot before running - shows raw data only
plot(obj, drawPoints = TRUE)
```

After running the model, the plot includes additional components and forecasts:

```{r run-before-plot}
obj$run()

# Plot after running - shows SA, trend, and forecast
plot(obj, drawPoints = TRUE, showOutliers = FALSE)
```

# Seasonal and Irregular Components

Visualize the seasonal and irregular components with `plotSeasIrrCal()`:

```{r plot-seas-irr}
plotSeasIrrCal(obj)
```

# Residual Analysis

`plotResiduals()` creates interactive diagnostic plots using `plotly`. You can examine autocorrelation (ACF) or partial autocorrelation (PACF):

```{r plot-residuals}
# ACF of residuals
plotResiduals(obj, which = "acf")

# PACF of residuals
plotResiduals(obj, which = "pacf")
```

# Highcharter Alternative

For alternative styling, use `hchart()` which leverages the `highcharter` package:

```{r hchart-plot}
hchart(obj)
```

# Plotly Integration

Some methods return `plotly` objects directly, allowing further customization:

```{r plotly-custom}
# ACF plot with plotly
p <- plotResiduals(obj, which = "acf")
p  # The plotly object can be further customized
```

# Repository

Source code and issue tracker: [https://github.com/statistikat/persephone3](https://github.com/statistikat/persephone3)
