---
title: "Get started with persephone3"
description: >
  A comprehensive introduction to seasonal adjustment with persephone3,
  covering installation, basic usage, and parameter customization.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with persephone3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{persephone3}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/persephone-",
  out.width = "100%"
)
```

# Introduction

`{persephone3}` provides an object-oriented wrapper around the seasonal adjustment packages in the [rjdverse](https://github.com/rjdverse/), specifically `{rjd3x13}` and `{rjd3tramoseats}`. The package performs time series adjustments using the Java library [JDemetra+](https://jdemetra-new-documentation.netlify.app/) with a focus on batch processing, hierarchical time series, and interactive visualizations.

## Installation

Install `{persephone3}` and its dependencies from GitHub:

```r
# Install rjdverse dependencies first
remotes::install_github("rjdverse/rjd3toolkit@*release")
remotes::install_github("rjdverse/rjd3x13@*release")
remotes::install_github("rjdverse/rjd3tramoseats@*release")

# Install persephone3
remotes::install_github("statistikat/persephone3")
```

# Basic Workflow

## Creating a Seasonal Adjustment Object

The first step is to create a seasonal adjustment object using either `perX13()` for X-13ARIMA-SEATS or `perTramo()` for TRAMO/SEATS:

```{r setup}
library(persephone3)
```

```{r create-object}
# X-13ARIMA-SEATS with default rsa3 template
obj_x13 <- perX13(AirPassengers)
obj_x13

# TRAMO/SEATS with default template
obj_tramo <- perTramo(AirPassengers)
obj_tramo
```

The `template` argument determines the default specification. Available templates differ between X-13 and TRAMO/SEATS:

### X-13ARIMA-SEATS Templates (`perX13()`)

| Template | Log/Level Detection | Outlier Detection | Calendar Effects | ARIMA Model |
|----------|---------------------|-------------------|------------------|-------------|
| `rsa0` | None | None | None | Airline (+mean) |
| `rsa1` | Automatic | AO, LS, TC | None | Airline (+mean) |
| `rsa2c` | Automatic | AO, LS, TC | 2 TD vars + Easter | Airline (+mean) |
| `rsa3` | Automatic | AO, LS, TC | None | Automatic |
| `rsa4` | Automatic | AO, LS, TC | 2 TD vars + Easter | Automatic |
| `rsa5c` | Automatic | AO, LS, TC | 7 TD vars + Easter | Automatic |

**Default**: `"rsa3"`

See [`rjd3x13::x13_spec()`](https://rjdverse.github.io/rjd3x13/reference/x13_spec.html) for details.

### TRAMO/SEATS Templates (`perTramo()`)

| Template | Log/Level Detection | Outlier Detection | Calendar Effects | ARIMA Model |
|----------|---------------------|-------------------|------------------|-------------|
| `rsa0` | None | None | None | Airline (+mean) |
| `rsa1` | Automatic | AO, LS, TC | None | Airline (+mean) |
| `rsa2` | Automatic | AO, LS, TC | 2 TD vars + Easter | Airline (+mean) |
| `rsa3` | Automatic | AO, LS, TC | None | Automatic |
| `rsa4` | Automatic | AO, LS, TC | 2 TD vars + Easter | Automatic |
| `rsa5` | Automatic | AO, LS, TC | 7 TD vars + Easter | Automatic |
| `rsafull` | Automatic | AO, LS, TC | Automatic | Automatic |

**Default**: `"rsa3"` for `perTramo()`, `"rsa5"` for `tramoseats_spec()`

See [`rjd3tramoseats::tramoseats_spec()`](https://rjdverse.github.io/rjd3tramoseats/reference/tramoseats_spec.html) for details.

**Abbreviations**:
- **AO**: Additive Outlier
- **LS**: Level Shift
- **TC**: Temporary Change
- **TD**: Trading Day

## Running the Model

Execute the seasonal adjustment with the `run()` method:

```{r run-model}
obj_x13$run()
```

After running, you can access the results through various methods.

## Inspecting Results

### Print and rjd3 SA-Output

```{r inspect-results}
# Print object information
obj_x13

# Access the rjd3 output object
obj_x13$output
```

### Diagnostics

```{r diagnostics}
# Get diagnostic statistics
obj_x13$diagnosticsTable()
```

### Visualization

`persephone3` provides multiple visualization options:

```{r plot-method, eval = FALSE}
# dygraphs plot method
obj_x13$plot()

# Highcharter interactive plot
hchart(obj_x13)
```

## Other Available Plots

Additional visualization methods are available:

- `obj$plotSpectrum()`: Spectral plots
- `obj$plotSeasIrrCal()`: Seasonal, irregular, and calendar effects
- `obj$plotResiduals()`: Residual diagnostics (ACF, PACF, QQ, histogram)
- `obj$plotSiRatios()`: Seasonal-to-Irregular ratio plot

For detailed documentation on all plotting options, see `vignette("persephone-plotting")`.

# Customizing Parameters

## Parameter Structure

`persephone3` provides a unified interface to the underlying `{rjd3toolkit}` and `{rjd3x13}` specification functions. Parameters are organized into **spec-blocks**, where each block corresponds to a specific `{rjd3*}` function.

The parameter name format is:

```
block.parameter
```

- **block**: The first part (before the dot) indicates the spec-block and maps to an `{rjd3*}` function
- **parameter**: The second part (after the dot) specifies the parameter within that block

When you pass parameters to `updateParams()` or during object creation, `persephone3` internally routes them to the appropriate `{rjd3*}` function based on the block name.

## Updating Parameters

Use `updateParams()` to modify seasonal adjustment settings. Parameters are passed directly and mapped to the corresponding `{rjd3*}` functions:

```{r update-params}
# Disable log transformation and enable iterative estimation
obj_x13$updateParams(
  transform.fun = "None",
  arima.p = 1,
  arima.d = 2,
  arima.q = 0,
  arima.bp = 1,
  arima.bd = 1,
  arima.bq = 0
)

# Re-run with new parameters
obj_x13$run()

# Check updated diagnostics
obj_x13$diagnosticsTable()
```

## Available Spec-Blocks

Each spec-block corresponds to a specific function in `{rjd3toolkit}` or `{rjd3x13}`. You can pass parameters directly using the block.parameter notation, and `persephone3` will route them to the appropriate `{rjd3*}` function:

| Block | `{rjd3*}` Function | Package | Description |
|-------|----------|---------|-------------|
| `arima` | [`set_arima()`](https://rjdverse.github.io/rjd3toolkit/reference/set_arima.html) | `{rjd3toolkit}` | ARIMA model specification |
| `automdl` | [`set_automodel()`](https://rjdverse.github.io/rjd3toolkit/reference/set_automodel.html) | `{rjd3toolkit}` | Automatic model selection |
| `basic` | [`set_basic()`](https://rjdverse.github.io/rjd3toolkit/reference/set_basic.html) | `{rjd3toolkit}` | Basic settings (series type, frequency, etc.) |
| `easter` | [`set_easter()`](https://rjdverse.github.io/rjd3toolkit/reference/set_easter.html) | `{rjd3toolkit}` | Easter effect adjustment |
| `estimate` | [`set_estimate()`](https://rjdverse.github.io/rjd3toolkit/reference/set_estimate.html) | `{rjd3toolkit}` | Estimation settings |
| `setout` | [`set_outlier()`](https://rjdverse.github.io/rjd3toolkit/reference/set_outlier.html) | `{rjd3toolkit}` | Outlier detection settings |
| `td` | [`set_tradingdays()`](https://rjdverse.github.io/rjd3toolkit/reference/set_tradingdays.html) | `{rjd3toolkit}` | Trading day effects |
| `transform` | [`set_transform()`](https://rjdverse.github.io/rjd3toolkit/reference/set_transform.html) | `{rjd3toolkit}` | Log/Box-Cox transformation |
| `addout` | [`add_outlier()`](https://rjdverse.github.io/rjd3toolkit/reference/add_outlier.html) | `{rjd3toolkit}` | Add specific outliers |
| `remout` | [`remove_outlier()`](https://rjdverse.github.io/rjd3toolkit/reference/remove_outlier.html) | `{rjd3toolkit}` | Remove outliers |
| `addramp` | [`add_ramp()`](https://rjdverse.github.io/rjd3toolkit/reference/add_ramp.html) | `{rjd3toolkit}` | Add ramp adjustments |
| `remramp` | [`remove_ramp()`](https://rjdverse.github.io/rjd3toolkit/reference/remove_ramp.html) | `{rjd3toolkit}` | Remove ramp adjustments |
| `usrdefvar` | [`add_usrdefvar()`](https://rjdverse.github.io/rjd3toolkit/reference/add_usrdefvar.html) | `{rjd3toolkit}` | Add user-defined variables |
| `x11` | [`set_x11()`](https://rjdverse.github.io/rjd3x13/reference/set_x11.html) | `{rjd3x13}` | X-11 specific options (X-13 only) |
| `benchmarking` | [`set_benchmarking()`](https://rjdverse.github.io/rjd3toolkit/reference/set_benchmarking.html) | `{rjd3toolkit}` | Benchmarking settings |

**Note:** The parameter names you use in `persephone3` (e.g., `arima.p`, `transform.fun`) correspond to the argument names in the underlying `{rjd3*}` functions. For complete parameter documentation, refer to the linked `{rjd3*}` function documentation.


## Simplified Syntax for Some parameters

For certain parameters (e.g., outliers), `persephone3` provides a simplified syntax.
Instead of using `addout.type = c("AO","LS")` and `addout.date = c("1953-07-01","1954-06-01")`, you can write:

```{r simplified-outliers, eval = FALSE}
obj_x13$updateParams(
  outliers = c("AO1953.7", "LS1954.6")
)
obj_x13$run()
```

This combines the outlier type and date into a single, more intuitive format.

# Accessing Specific Results

## Preprocessing Results

```{r preprocessing}
# Access preprocessing output
obj_x13$output$preprocessing
```

## Original Series

```{r orig-series}
# Original series
orig_series <- obj_x13$ts
head(orig_series)
```

## Seasonally Adjusted Series

```{r sa-series}
# Get seasonally adjusted series
sa_series <- obj_x13$adjusted
head(sa_series)
```

## Access Original, Seasonally Adjusted and Trend Series via `{rjd3*}` `user_defined` Slot

```{r rjd3-user_defined}
# Original series
head(obj_x13$output$user_defined$y)
# Seasonally adjusted
head(obj_x13$output$user_defined$sa)
# Trend
head(obj_x13$output$user_defined$t)
```

# Next Steps

- See `vignette("persephone-plotting")` for detailed visualization options
- See `vignette("persephone-hierarchical")` for hierarchical time series adjustment
- Check `?perX13` and `?perTramo` for complete parameter documentation
- Explore `{rjd3toolkit}` and `{rjd3x13}` documentation for advanced spec-block options

# Common Issues

## Java not found

If you encounter Java-related errors, ensure Java 17+ is installed:

```r
# Check Java installation
Sys.getenv("JAVA_HOME")
```

## Model convergence issues

If the model fails to converge:

1. Try a simpler template (e.g., `"rsa0"` instead of `"rsa3"`)
2. Disable automatic outlier detection
3. Check for extreme values or missing data in your series

# Repository

Source code and issue tracker: [https://github.com/statistikat/persephone3](https://github.com/statistikat/persephone3)
