---
title: "Production-ready seasonal adjustment workflows"
description: >
  A narrative version of the NTTS Persephone slides, updated for
  `persephone3`.
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Production-ready seasonal adjustment workflows}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

This vignette turns the original NTTS slide deck in
`vignettes/resources/slides_ntts.Rmd.src` into a compact workflow article for
the current `persephone3` package. The goal is the same as in the slides:
build production-ready seasonal-adjustment pipelines in R, while staying close
to the underlying JDemetra+ ecosystem.

`persephone3` wraps `rjd3x13` and `rjd3tramoseats` in a small set of object
types that make routine production work easier:

```{r workflow-table}
workflow_map <- data.frame(
  constructor = c("perX13() / perTramo()", "perBatch()", "perHts()"),
  role = c(
    "Create a single seasonal-adjustment object.",
    "Run the same workflow over a collection of aligned series.",
    "Build direct and indirect adjustments for hierarchical aggregates."
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  workflow_map,
  col.names = c("Constructor", "Role"),
  caption = "Core object types in persephone3"
)
```

## A single series

We start with the simplest workflow: define one X-13ARIMA-SEATS model, update
its specification, and run it.

```{r single-series}
library(persephone3)

data(AirPassengers, package = "datasets")

obj_x13 <- perX13(AirPassengers, template = "rsa3")
obj_x13$updateParams(transform.fun = "Log")
obj_x13$run()
```

The resulting object keeps the original series, the adjusted series, and the
full model specification. For a quick check, we can compare the original and
adjusted paths directly.

```{r single-plot}
single_plot <- data.frame(
  date = zoo::as.Date(zoo::as.yearmon(stats::time(obj_x13$ts))),
  original = as.numeric(obj_x13$ts),
  adjusted = as.numeric(obj_x13$adjusted)
)

ggplot2::ggplot(single_plot, ggplot2::aes(date)) +
  ggplot2::geom_line(
    ggplot2::aes(y = original, color = "Original"),
    linewidth = 0.5,
    alpha = 0.7
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = adjusted, color = "Adjusted"),
    linewidth = 0.6
  ) +
  ggplot2::scale_color_manual(
    values = c("Original" = "#6c757d", "Adjusted" = "#005f73")
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Passengers",
    color = NULL,
    title = "AirPassengers before and after seasonal adjustment"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

## A small batch

Production work usually means many related series rather than one. The next
example takes three monthly production-index series, creates one
`perX13()` object for each of them, and processes them as a batch.

```{r prepare-panel}
data(pi_caladj)
data(weights_pi_ea19)
data(weights_pi_eu28)

pi_panel <- stats::window(pi_caladj[, -c(1, 2)], end = c(2019, 5))

panel_to_persephone <- function(x, template = "rsa3") {
  out <- lapply(as.data.frame(x), function(series) {
    perX13(
      stats::ts(
        series,
        start = stats::start(x),
        frequency = stats::frequency(x)
      ),
      template = template
    )
  })

  names(out) <- colnames(x)
  out
}

ts_28 <- panel_to_persephone(pi_panel)
```

```{r batch}
batch_fit <- perBatch(list = ts_28[c("AT", "DE", "FR")])
batch_fit$run()

batch_summary <- data.frame(
  component = names(batch_fit$components),
  start = vapply(
    batch_fit$components,
    function(x) paste(stats::start(x$adjusted), collapse = "-"),
    character(1)
  ),
  end = vapply(
    batch_fit$components,
    function(x) paste(stats::end(x$adjusted), collapse = "-"),
    character(1)
  ),
  transform = vapply(
    batch_fit$components,
    function(x) x$params$regarima$transform$fn,
    character(1)
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  batch_summary,
  caption = "A small batch of aligned series"
)
```

This pattern is useful when the series are processed together but do not belong
to a formal hierarchy.

## A weighted hierarchy

The main production use case in the NTTS slides was a hierarchy of European
production indices. We can recreate that workflow with `perHts()`. Here we use
EA-19 as an intermediate aggregate and retain the EU-28 weights for the
aggregation steps.

```{r hierarchy}
ea19 <- weights_pi_ea19$country
non_ea19 <- weights_pi_eu28$country[!weights_pi_eu28$country %in% ea19]

w_ea19 <- weights_pi_eu28$weight[weights_pi_eu28$country %in% ea19]
w_non_ea19 <- weights_pi_eu28$weight[!weights_pi_eu28$country %in% ea19]

hts_ea19 <- perHts(
  list = ts_28[ea19],
  method = "x13",
  weights = w_ea19
)

hts_non_ea19 <- perHts(
  list = ts_28[non_ea19],
  weights = w_non_ea19
)

hts_eu28 <- perHts(
  EA19 = hts_ea19,
  non_EA19 = hts_non_ea19
)

hts_eu28$run()
```

Once the hierarchy has been run, the root object contains both the direct and
indirect adjustment of the aggregate.

```{r hierarchy-plot}
aggregate_plot <- data.frame(
  date = zoo::as.Date(zoo::as.yearmon(stats::time(hts_eu28$ts))),
  original = as.numeric(hts_eu28$ts),
  direct = as.numeric(hts_eu28$adjustedDirect),
  indirect = as.numeric(hts_eu28$adjustedIndirect)
)

ggplot2::ggplot(aggregate_plot, ggplot2::aes(date)) +
  ggplot2::geom_line(
    ggplot2::aes(y = original, color = "Original aggregate"),
    linewidth = 0.5,
    alpha = 0.65
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = direct, color = "Direct adjustment"),
    linewidth = 0.6
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = indirect, color = "Indirect adjustment"),
    linewidth = 0.6,
    linetype = "22"
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "Original aggregate" = "#6c757d",
      "Direct adjustment" = "#0a9396",
      "Indirect adjustment" = "#bb3e03"
    )
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Index",
    color = NULL,
    title = "Weighted EU-28 aggregate: direct and indirect adjustment"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

The hierarchy also feeds directly into the Eurostat-style quality report table.

```{r qr-table}
qr_table <- generateQrTable(hts_eu28)

knitr::kable(
  head(qr_table, n = 4),
  caption = "First rows of the quality-report table"
)
```

## Stabilizing a production specification

Two helpers from the original slides are especially useful in recurring
production runs:

- `fixModel()` freezes the currently selected ARIMA model.
- `fixOutlier()` can move detected outliers into fixed regressors while leaving
  the trailing part of the sample open for fresh detection.

The first one is easy to inspect directly. After `fixModel()`, automatic model
selection is disabled in the stored specification.

```{r fix-model}
obj_fixed <- perX13(AirPassengers, template = "rsa3")
obj_fixed$run()
obj_fixed$fixModel()

data.frame(
  automodel_enabled = obj_fixed$params$regarima$automodel$enabled,
  regular_d = obj_fixed$params$regarima$arima$d,
  seasonal_d = obj_fixed$params$regarima$arima$bd
)
```

The same helpers also recurse through `perBatch()` and `perHts()` objects.

When the workflow already carries a user-defined outlier block, `fixOutlier()`
can be used in the same spirit:

```{r fix-outlier, eval = FALSE}
obj_fixed$fixOutlier(timespan = 12)
obj_fixed$run()
```

## Country-specific trading-day regressors

The slide deck ended with a `genTd()` example for holiday-aware trading-day
regressors. The same idea is still useful when the default calendars do not
match a country's production pattern.

```{r gentd}
hdAT <- genTd(
  firstYear = 2000,
  lastYear = 2026,
  hd = list(
    "01-01", "01-06", "05-01", "easter+1", "easter+39",
    "easter+50", "easter+60", "08-15", "10-26", "11-01",
    "12-08", "12-24", "12-25", "12-26", "12-31"
  ),
  weight = c(rep(1, 11), 0.5, rep(1, 2), 0.5)
)

knitr::kable(
  round(head(hdAT[[3]][, c("Monday", "lpYear")]), 3),
  caption = "Centered holiday-adjusted trading-day regressors"
)
```

Those regressors can then be passed into a model specification as user-defined
calendar variables:

```{r gentd-model, eval = FALSE}
obj_td <- perX13(
  AirPassengers,
  template = "rsa3",
  context = list(calendar_td = hdAT[[3]]),
  td.option = "UserDefined",
  td.uservariable = "calendar_td"
)

obj_td$run()
```

## Closing remarks

The original NTTS slides framed `persephone` as an infrastructure layer on top
of JDemetra+. That description still fits `persephone3`: single series, batches
of aligned series, weighted hierarchies, production freezing with
`fixModel()`, and custom calendar regressors all live in the same object model,
which makes it much easier to move from exploratory work to repeatable
production runs.
